Normalization by Evaluation for Sized Dependent Types

### 2017-02-21 nbe/icfp17.tex
[ ICFP 17 ] Properties of shape (mostly failing ones).

Judgements.
\[
\begin{array}{l@{\qquad}l}
  \der \Gamma & \mbox{context $\Gamma$ is well-formed} \\
  \Gamma(i) = \eann T & \mbox{in context $\Gamma$, de Bruijn index $i$ has type $T$ and annotation $\evar$} \\
  \Gamma \der a : \Size & \mbox{in context $\Gamma$, size expression $a$ is well-formed} \\
  \Gamma \der t : T & \mbox{in context $\Gamma$, term $t$ has type $T$} \\
  \Gamma \der t = t' : T & \mbox{in context $\Gamma$, terms $t$ and $t'$ are equal of type $T$} \\
%  \Gamma \mid U \der e : T & \mbox{in context $\Gamma$, elimination $e$ goes from $U$ to $T$} \\
%  \Gamma \mid U \der e = e' : T & \mbox{in context $\Gamma$, eliminations $e$ and $e'$ are equal} \\
  \Gamma \der T \leq T' & \mbox{in context $\Gamma$, type $T$ is a subtype of $T'$} \\
  \suty \Gamma \sigma \tau \Delta & \mbox{$\sigma$/$\tau$ is a valid term/type-level substitution for $\Delta$} \\
  \sueq \Gamma \sigma {\sigma'} \tau \Delta & \mbox{$\sigma$/$\sigma'$/$\tau$ are a equal term/term/type-level substitutions for $\Delta$} \\
\end{array}
\]
Derived judgements.
\[
\begin{array}{lll@{~}l@{\qquad}l}
  \Gamma \der T & \defas & \Gamma \der T : s & \mforsome s
    & \mbox{in context $\Gamma$, type $T$ is well-formed} \\
  \Gamma \der T = T' & \defas & \Gamma \der T = T' : s & \mforsome s
    & \mbox{in context $\Gamma$, types $T$ and $T'$ are equal} \\
  \Gamma \der a \leq b : \Size & \defas & \multicolumn 2 {l@{\qquad}} {\Gamma \der a : \Size \mand \Gamma \der b : \Size \mand a \leq b}
    & \mbox{sizes $a$ and $b$ are well-formed and comparable} \\
%    & \mbox{in context $\Gamma$, well-formed size $a$ is less or equal than well-formed size $b$} \\
\end{array}
\]
Inference rules.

@@

\fbox{$\Gamma \der t : T$}  (implies $\der \Gamma$ and $\resurrect \Gamma \der T$ [and $\resurrect \Gamma \der t : T$]. Note: no rule for $\Gamma \der \Size : s$.)
\begin{gather*}
  \ru{\der \Gamma
    }{\Gamma \der \Set[\ell] : \Set[\ell']
    }{\ell {<} \ell'}
\qquad
  \ru{\Gamma \der U : s \qquad
      \cext \Gamma U \der T : s
    }{\Gamma \der \piT U T : s}
\qquad
  \ru{\cext \Gamma \Size \der T : s
    }{\Gamma \der \epiT \Size T : s}
\qquad
  \ru{%\der \Gamma \qquad
      \Gamma \der a : \Size
    }{\Gamma \der \Nat a : s}
\\[2ex]
  \rux{\der \Gamma \qquad \Gamma(i) = \erann[\noterased] T
    }{\Gamma \der \ind i : T
    }{T \not= \Size}
\qquad
  \ru{\eext \Gamma U \der t : T
    }{\Gamma \der \lambda t : \epiT U T}
\qquad
  \ru{\Gamma \der t : \piT U T \qquad
      \Gamma \der u : U
    }{\Gamma \der t\,u : T[u]}
% \qquad
%   \ru{\Gamma \der t : T \qquad
%       \Gamma \mid T \der e : T'
%     }{\Gamma \der t\,e : T'}
\qquad
  \ru{\Gamma \der t : \piT \Size T \qquad
      \Gamma \der a : \Size
    }{\Gamma \der t\,a : T[a]}
\\[2ex] % \qquad
  \ru{\Gamma \der a : \Size \qquad
      \resurrect\Gamma \der b : \Size
    }{\Gamma \der \zero{a} : \Nat (b + 1)}
\qquad
  \ru{\Gamma \der a : \Size \qquad
      \Gamma \der t : \Nat b
    }{\Gamma \der \suc a t : \Nat (b + 1)}
\qquad
  \ru{\Gamma \der t : \forallT T \qquad
      \Gamma \der a : \Size \qquad
      \resurrect\Gamma \der b : \Size
    }{\Gamma \der t \ann a : T[b]}
\\[2ex]
  \ru{\Gamma \der u : \Nat (a + 1) \qquad
      \Gamma \der T : \Nat (a + 1) \to \Set[\ell] \qquad
      \Gamma \der t_z : T \, (\zero{a}) \qquad
      \Gamma \der t_s : (t : \Nat a) \to T \, (\suc{a} t)
    }{\Gamma \der u\,\case[\ell] T {t_z} t_s : T \, u}
\\[2ex]
  \ru{\Gamma \der u : \Nat a \qquad
      \Gamma \der T : \FixK\,\ell \qquad %% \funT i \Size \Nat i \to \Set[\ell] \\
      \Gamma \der t : \FixT\, T
     }{\Gamma \der u\,\fix[\ell] T t : T \,a \, u}
\qquad
  \ru{\Gamma \der t : T \qquad
      \resurrect\Gamma \der T \leq T'
     }{\Gamma \der t : T'}
\end{gather*}
Note: Size is relevant in $T$ as we would like for instance $T\,i\,x = \Nat i$.
% \fbox{$\Gamma \mid U \der e : T$}
% \begin{gather*}
%   \ru{\cext \Gamma U \der T \qquad \Gamma \der u : U
%     }{\Gamma \mid \piT U T \der u : T[u]}
% \qquad
%   \ru{\cext \Gamma \Size \der T \qquad \Gamma \der a : \Size
%     }{\Gamma \mid \piT \Size T \der a : T[a]}
% \qquad
%   \ru{\cext \Gamma \Size \der T \qquad \Gamma \der a : \Size
%     }{\Gamma \mid \forallT T \der \ann a : T[a]}
% \\
%   \ru{
%     }{\Gamma \mid \Nat a \der \fix T t : T \ann a n}
% \end{gather*}

@@

Judgements.
\[
\begin{array}{l@{\qquad}l}
   \R k d  v     & \mbox{} \\
   \RTy k T  V  & \mbox{} \\
   \RNat k t  v & \mbox{} \\
   \RNe k t  m  & \mbox{} \\
\end{array}
\]
\fbox{$\R k d v$}
\begin{gather*}
  \ru{U \evalsto s \qquad
      \RTy k T V
    }{\R k {\down{U} T} V}
\qquad
  \ru{U \evalsto \Nat \alpha \qquad
      \RNat k u v
    }{\R k {\down{U} u} v}
\qquad
  \ru{U \evalsto N \qquad
      u \evalsto \up{T} n \qquad
      \RNe k n m
    }{\R k {\down{U} u} m}
\\[2ex]
  \ru{U \evalsto \piT A B \qquad
      \R{k+1} {\down{B[\up A \lev k]} (f\,\up A \lev k)} v
    }{\R k {\down{U} f} {\lambda v}}
\\[2ex]
  \ru{U \evalsto \piT \Size B \qquad
      \R{k+1} {\down{B[\lev k]} (f\,\lev k)} v
    }{\R k {\down{U} f} {\lambda v}}
\qquad
  \ru{U \evalsto \forallT B \qquad
      \R{k+1} {\down{B[\lev k]} (f\,\ann{\lev k})} v
    }{\R k {\down{U} f} {\lambda v}}
\end{gather*}

\fbox{$\RNat k t v $}
\begin{gather*}
  \ru{t \evalsto \up {\Nat \alpha} n \qquad
      \RNe k n m
    }{\RNat k t m}
\qquad
  \ru{t \evalsto \zero{\alpha} \qquad
      \RSize k {\alpha} a
    }{\RNat k t {\zero a}}
\qquad
  \ru{t \evalsto \suc{\alpha}u \qquad
      \RSize k {\alpha} a \qquad
      \RNat k u v
    }{\RNat k t {\suc a v}}
\end{gather*}

@@

\fbox{$\RNe k n m$} and \fbox{$\RElim k e e^v$}
\begin{gather*}
%   \ru{t \evalsto \lev j
%     }{\RNe k t {\ind{k-1-j}}}
% \qquad
  \ru{%t \evalsto \lev j\,\vec e \qquad
      \RElim k {e_i} {e^v_i} \mforall i
    }{\RNe k {\lev j\,\vec e} {\ind{k \monus (1+j)}\,\vec e^v}}
\qquad
% \end{gather*}
% \begin{gather*}
  \ru{\R k d v
    }{\RElim k d v}
\qquad
  \ru{\RSize k \alpha b
    }{\RElim k \alpha b}
\qquad
  \ru{\RSize k \alpha b
    }{\RElim k {\ann \alpha} {\ann b}}
\\[2ex]
  \ru{\RTy k T V \qquad
      \R k {d_z} {v_z} \qquad
      \R k {d_s} {v_s}
    }{\RElim k {(\case[\ell] T {d_z} {d_s})} {\case[\ell] V {v_z} {v_s}}}
\qquad
  \ru{\RTy k T V \qquad
      \R k d v
    }{\RElim k {(\fix[\ell] T d)} {\fix[\ell] V v}}
\end{gather*}
\fbox{$\RTy k T V$}
\begin{gather*}
  \ru{T \evalsto \Set[\ell]
    }{\RTy k T {\Set[\ell]}}
\qquad
  \ru{T \evalsto \Nat \alpha \qquad
      \RSize k \alpha b
    }{\RTy k T {\Nat b}}
\qquad
  \ru{T \evalsto \up{U}n \qquad
      \RNe k n m
    }{\RTy k T m}
\\[2ex]
  \ru{T \evalsto \piT A B \qquad
      \RTy k A {V_a} \qquad
      \RTy {k+1} B {V_b}
    }{\RTy k T {\piT {V_a} V_b}}
\qquad
  \ru{T \evalsto \epiT \Size B \qquad
      \RTy {k+1} B {V_b}
    }{\RTy k T {\epiT \Size V_b}}
\end{gather*}

@@

\[
\fbox{$\NEE$} \defas \{ (\up T n, \up{T'}{n'})
  \msuchthat T \evalsto N \mand T' \evalsto N' \mand n = n' \in \NE \}
\]
\fbox{$\NAT(\alpha)$}
\begin{gather*}
  \ru{t \evalsto \zero \alpha \qquad
      t' \evalsto \zero{\alpha'}
    }{t = t' \in \NAT(\beta+1)}
\qquad
  \ru{t \evalsto \suc \alpha u \qquad
      t' \evalsto \suc {\alpha'} {u'} \qquad
      u = u' \in \NAT(\beta)
    }{t = t' \in \NAT(\beta+1)}
\\[2ex]
  \ru{t \evalsto \up{\Nat \alpha} n \qquad
      t' \evalsto \up{\Nat \alpha'} n' \qquad
      n = n' \in \NE
    }{t = t' \in \NAT(\beta)}
\end{gather*}

@@

\begin{lemma}[PER Subsumption]
\label{lem:subsump} \hfill
\begin{enumerate}
\item If $\alpha \leq \beta$ then $\NAT(\alpha) \subseteq \NAT(\beta)$.
\item If $\F(\alpha) \leq \F'(\alpha)$ for all $\alpha \in \SIZE$, then $\FORALL \F \subseteq \FORALL \F'$.
\item IF $\A' \subseteq \A$ and $\F(u) \subseteq \F'(u)$ for all $u \in \A'$, then $\PIAF \subseteq \PI\,\A'\,\F'$.
\end{enumerate}
\end{lemma}

@@

By induction on $\ell \in \NN$ we define the PER family
$\_ = \_ \in \Set[\ell]$ of types
together with the extension $\EL[\ell] T$ (for $T = T' \in \Set[\ell]$)
which is PER of values of type $A$.
The rules for \fbox{$T = T' \in \SET[\ell]$} are:
\[
\begin{array}{l@{\qquad}l}
  \ru{T \evalsto N \qquad T' \evalsto N'
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \NEE
\\[3ex]
  \ru{T \evalsto \Nat\,\alpha \qquad T' \evalsto \Nat\,\alpha
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \NAT(\alpha)
\\[3ex]
  \rux{T \evalsto \Set[\ell'] \qquad T' \evalsto \Set[\ell']
     }{T = T' \in \SET[\ell]
     }{\ell'{<}\ell}
  & \EL[\ell](T) = \SET[\ell']
\\[3ex]
  \rul{T \evalsto \piT A B \qquad
      T' \evalsto \piT{A'}{B'} \\
      A = A' \in \SET[\ell] \qquad
      B[u] = B'[u'] \in \SET[\ell] \mforall u = u' \in \EL[\ell](A)
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \PI(\EL[\ell](A),\;u \mapsto \EL[\ell](B[u]))
\\[3ex]
  \rul{T \evalsto \piT \Size B \quad
      T' \evalsto \piT{\Size}{B'} \quad
      B[\alpha] = B'[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \PI(\SIZE,\; \alpha \mapsto \EL[\ell](B[\alpha]))
\\[3ex]
  \rul{T \evalsto \forallT B \quad
      T' \evalsto \forallT{B'} \quad
      B[\alpha] = B'[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \FORALL(\alpha \mapsto \EL[\ell](B[\alpha]))
\end{array}
\]
All relations involved here are closed under weak head expansion.
\begin{lemma}[Well-definedness]
  Let $\DD :: T_1 = T_2 \in \SET[\ell]$.
  \begin{enumerate}
  \item Symmetry: $T_2 = T_1 \in \SET[\ell]$.
  \item Transitivity: If $T_2 = T_3 \in \SET[\ell]$ then $T_1 = T_3 \in \SET[\ell]$.
  \item Extension is PER: $\EL[\ell](T_1) = \EL[\ell](T_2)$ and ``both'' are PERs.
  \end{enumerate}
\end{lemma}
\begin{proof}
  Simultaneously by induction on $\DD$.
\end{proof}

@@

\begin{lemma}[Universe subsumption]
  \label{lem:usubsump}
  If $\ell \leq \ell'$ then $\SET[\ell] \subseteq \SET[\ell']$.
\end{lemma}
\begin{proof}
  Prove $T = T' \in \SET[\ell']$ by induction on $T = T' \in \SET[\ell]$.
  For the base types this is direct, let us look at the function space.
\[
  \rul{T \evalsto \piT A B \qquad
      T' \evalsto \piT{A'}{B'} \qquad
      A = A' \in \SET[\ell] \qquad
      B[u] = B'[u'] \in \SET[\ell] \mforall u = u' \in \EL[\ell](A)
    }{T = T' \in \SET[\ell]}
\]
  By induction hypothesis, $A = A' \in \SET[\ell']$, and we have $\EL[\ell'](A) = \EL[\ell](A)$ by Lemma~\ref{lem:indep}.  Assuming $u = u' \in \EL(A)$, we get $B[u] = B'[u'] \in \SET[\ell']$ by induction hypothesis on $B[u] = B'[u'] \in \SET[\ell]$.
\end{proof}

\subsection{Type shapes}

>>>>

Judgements.
\[
\begin{array}{l@{\qquad}l}
  \der \Gamma & \mbox{context $\Gamma$ is well-formed} \\
  \Gamma(i) = \eann T & \mbox{in context $\Gamma$, de Bruijn index $i$ has type $T$ and annotation $\evar$} \\
  \Gamma \der a : \Size & \mbox{in context $\Gamma$, size expression $a$ is well-formed} \\
  \Gamma \der t : T & \mbox{in context $\Gamma$, term $t$ has type $T$} \\
  \Gamma \der t = t' : T & \mbox{in context $\Gamma$, terms $t$ and $t'$ are equal of type $T$} \\
%  \Gamma \mid U \der e : T & \mbox{in context $\Gamma$, elimination $e$ goes from $U$ to $T$} \\
%  \Gamma \mid U \der e = e' : T & \mbox{in context $\Gamma$, eliminations $e$ and $e'$ are equal} \\
  \Gamma \der T \leq T' & \mbox{in context $\Gamma$, type $T$ is a subtype of $T'$} \\
  \Gamma \der T : \Adm \ell & \mbox{in context $\Gamma$ type constructor $T$ is admissible for recursion} \\
  \suty \Gamma \sigma \tau \Delta & \mbox{$\sigma$/$\tau$ is a valid term/type-level substitution for $\Delta$} \\
  \sueq \Gamma \sigma {\sigma'} \tau \Delta & \mbox{$\sigma$/$\sigma'$/$\tau$ are a equal term/term/type-level substitutions for $\Delta$} \\
\end{array}
\]
Derived judgements.
\[
\begin{array}{lll@{~}l@{\qquad}l}
  \Gamma \der T & \defas & \Gamma \der T : s & \mforsome s
    & \mbox{in context $\Gamma$, type $T$ is well-formed} \\
  \Gamma \der T = T' & \defas & \Gamma \der T = T' : s & \mforsome s
    & \mbox{in context $\Gamma$, types $T$ and $T'$ are equal} \\
  \Gamma \der a = b : \Size & \defas & \multicolumn 2 {l@{\qquad}} {\Gamma \der a : \Size \mand a = b}
    & \mbox{sizes $a$ and $b$ are well-formed and identical} \\
  \Gamma \der a \leq b : \Size & \defas &
     \multicolumn 2 {l@{\qquad}} {\Gamma \der a : \Size \mand \Gamma \der b : \Size \mand a \leq b}
%     \multicolumn 2 {l@{\qquad}} {\Gamma \der a : \Size \mand \Gamma \der b : \Size \mand a \leq b}
    & \mbox{sizes $a$ and $b$ are well-formed and comparable} \\
%    & \mbox{in context $\Gamma$, well-formed size $a$ is less or equal than well-formed size $b$} \\
\end{array}
\]
Inference rules.

@@

\fbox{$\Gamma \der t : T$}  (implies $\der \Gamma$ and $\resurrect \Gamma \der T$ [and $\resurrect \Gamma \der t : T$]. Note: no rule for $\Gamma \der \Size : s$.)
\begin{gather*}
  \ru{\der \Gamma
    }{\Gamma \der \Set[\ell] : \Set[\ell']
    }{\ell {<} \ell'}
\qquad
  \ru{\Gamma \der U : s \qquad
      \cext \Gamma U \der T : s
    }{\Gamma \der \piT U T : s}
\qquad
  \ru{\cext \Gamma \Size \der T : s
    }{\Gamma \der \epiT \Size T : s}
\qquad
  \ru{%\der \Gamma \qquad
      \Gamma \der a : \Size
    }{\Gamma \der \Nat a : s}
\\[2ex]
  \rux{\der \Gamma \qquad \Gamma(i) = \erann[\noterased] T
    }{\Gamma \der \ind i : T
    }{T \not= \Size}
\qquad
  \ru{\eext \Gamma U \der t : T
    }{\Gamma \der \lambda t : \epiT U T}
\qquad
  \ru{\Gamma \der t : \piT U T \qquad
      \Gamma \der u : U
    }{\Gamma \der t\,u : T[u]}
% \qquad
%   \ru{\Gamma \der t : T \qquad
%       \Gamma \mid T \der e : T'
%     }{\Gamma \der t\,e : T'}
\qquad
  \ru{\Gamma \der t : \piT \Size T \qquad
      \Gamma \der a : \Size
    }{\Gamma \der t\,a : T[a]}
\\[2ex] % \qquad
  \ru{%\Gamma \der a : \Size \qquad
      \resurrect\Gamma \der a, b : \Size
    }{\Gamma \der \zero{a} : \Nat (b + 1)}
\qquad
  \ru{\resurrect\Gamma \der a : \Size \qquad
      \Gamma \der t : \Nat b
    }{\Gamma \der \suc a t : \Nat (b + 1)}
\qquad
  \ru{\Gamma \der t : \forallT T \qquad
      %\Gamma \der a : \Size \qquad
      \resurrect\Gamma \der a,b : \Size
    }{\Gamma \der t \ann a : T[b]}
\\[2ex]
  \ru{\Gamma \der u : \Nat (a + 1) \qquad
      \Gamma \der T : \Nat (a + 1) \to \Set[\ell] \qquad
      \Gamma \der t_z : T \, (\zero{a}) \qquad
      \Gamma \der t_s : (t : \Nat a) \to T \, (\suc{a} t)
    }{\Gamma \der u\,\case[\ell] T {t_z} t_s : T \, u}
\\[2ex]
  \ru{\Gamma \der u : \Nat a \qquad
      \Gamma \der T : \FixK\,\ell \qquad %% \funT i \Size \Nat i \to \Set[\ell] \\
      \Gamma \der t : \FixT\, T
     }{\Gamma \der u\,\fix[\ell] T t : T \,a \, u}
\qquad
  \ru{\Gamma \der t : T \qquad
      \resurrect\Gamma \der T \leq T'
     }{\Gamma \der t : T'}
\end{gather*}
Note: Size is relevant in $T$ as we would like for instance $T\,i\,x = \Nat i$.
% \fbox{$\Gamma \mid U \der e : T$}
% \begin{gather*}
%   \ru{\cext \Gamma U \der T \qquad \Gamma \der u : U
%     }{\Gamma \mid \piT U T \der u : T[u]}
% \qquad
%   \ru{\cext \Gamma \Size \der T \qquad \Gamma \der a : \Size
%     }{\Gamma \mid \piT \Size T \der a : T[a]}
% \qquad
%   \ru{\cext \Gamma \Size \der T \qquad \Gamma \der a : \Size
%     }{\Gamma \mid \forallT T \der \ann a : T[a]}
% \\
%   \ru{
%     }{\Gamma \mid \Nat a \der \fix T t : T \ann a n}
% \end{gather*}

@@

Judgements.
\[
\begin{array}{l@{\qquad}l}
   \R k d  v     & \mbox{} \\
   \RTy k T  V  & \mbox{} \\
   \RNat k t  v & \mbox{} \\
   \RNe k t  m  & \mbox{} \\
\end{array}
\]
\fbox{$\R k d v$}
\begin{gather*}
  \ru{U \evalsto s \qquad
      \RTy k T V
    }{\R k {\down{U} T} V}
\qquad
  \ru{U \evalsto \Nat \alpha \qquad
      \RNat k u v
    }{\R k {\down{U} u} v}
\qquad
  \ru{U \evalsto N \qquad
      \RUp k u m
      % u \evalsto \up{T} n \qquad \RNe k n m
    }{\R k {\down{U} u} m}
\\[2ex]
  \ru{U \evalsto \piT A B \qquad
      \R{k+1} {\down{B[\up A \lev k]} (f\,\up A \lev k)} v
    }{\R k {\down{U} f} {\lambda v}}
\\[2ex]
  \ru{U \evalsto \piT \Size B \qquad
      \R{k+1} {\down{B[\lev k]} (f\,\lev k)} v
    }{\R k {\down{U} f} {\lambda v}}
\qquad
  \ru{U \evalsto \forallT B \qquad
      \R{k+1} {\down{B[\lev k]} (f\,\ann{\lev k})} v
    }{\R k {\down{U} f} {\lambda v}}
\end{gather*}

\fbox{$\RUp k t m$}\ \  Read back neutrals under annotation, which is ignored.
\begin{gather*}
  \ru{t \evalsto \up T n \qquad
      \RNe k n m
    }{\RUp k t m}
\end{gather*}

\fbox{$\RNat k t v $}
\begin{gather*}
  \ru{% t \evalsto \up {\Nat \alpha} n \qquad
      %t \evalsto \up {T} n \qquad \RNe k n m
      \RUp k t m
    }{\RNat k t m}
\qquad
  \ru{t \evalsto \zero{\alpha} \qquad
      \RSize k {\alpha} a
    }{\RNat k t {\zero a}}
\qquad
  \ru{t \evalsto \suc{\alpha}u \qquad
      \RSize k {\alpha} a \qquad
      \RNat k u v
    }{\RNat k t {\suc a v}}
\end{gather*}

@@

\fbox{$\RNe k n m$} and \fbox{$\RElim k e e^v$}
\begin{gather*}
%   \ru{t \evalsto \lev j
%     }{\RNe k t {\ind{k-1-j}}}
% \qquad
  \ru{%t \evalsto \lev j\,\vec e \qquad
      \RElim k {e_i} {e^v_i} \mforall i
    }{\RNe k {\lev j\,\vec e} {\ind{k \monus (1+j)}\,\vec e^v}}
\qquad
% \end{gather*}
% \begin{gather*}
  \ru{\R k d v
    }{\RElim k d v}
\qquad
  \ru{\RSize k \alpha b
    }{\RElim k \alpha b}
\qquad
  \ru{\RSize k \alpha b
    }{\RElim k {\ann \alpha} {\ann b}}
\\[2ex]
  \ru{\RTy k T V \qquad
      \R k {d_z} {v_z} \qquad
      \R k {d_s} {v_s}
    }{\RElim k {(\case[\ell] T {d_z} {d_s})} {\case[\ell] V {v_z} {v_s}}}
\qquad
  \ru{\RTy k T V \qquad
      \R k d v
    }{\RElim k {(\fix[\ell] T d)} {\fix[\ell] V v}}
\end{gather*}
\fbox{$\RTy k T V$}
\begin{gather*}
  \ru{T \evalsto \Set[\ell]
    }{\RTy k T {\Set[\ell]}}
\qquad
  \ru{T \evalsto \Nat \alpha \qquad
      \RSize k \alpha b
    }{\RTy k T {\Nat b}}
\qquad
  \ru{% T \evalsto \up{U}n \qquad \RNe k n m
      \RUp k T m
    }{\RTy k T m}
\\[2ex]
  \ru{T \evalsto \piT A B \qquad
      \RTy k A {V_a} \qquad
      \RTy {k+1} B {V_b}
    }{\RTy k T {\piT {V_a} V_b}}
\qquad
  \ru{T \evalsto \epiT \Size B \qquad
      \RTy {k+1} B {V}
    }{\RTy k T {\epiT \Size V}}
\end{gather*}

@@

Neutrals in the value world $\D$.
\[
\fbox{$t = t' \in \NEE$} \defas
  t \evalsto \up T n \mand t' \evalsto \up{T'}{n'} \mand
  n = n' \in \NE
% \fbox{$\NEE$} \defas \{ (\up T n, \up{T'}{n'})
%   % \msuchthat T \evalsto N \mand T' \evalsto N' \mand
%   \mid n = n' \in \NE \}
\]
\fbox{$\NAT(\alpha)$}
\begin{gather*}
  \ru{% t \evalsto \up{\Nat \alpha} n \qquad
      % t' \evalsto \up{\Nat \alpha'} n' \qquad
      % n = n' \in \NE
      t = t' \in \NEE
    }{t = t' \in \NAT(\beta)}
\qquad
  \ru{t \evalsto \zero \alpha \qquad
      t' \evalsto \zero{\alpha'}
    }{t = t' \in \NAT(\beta+1)}
\qquad
  \ru{t \evalsto \suc \alpha u \qquad
      t' \evalsto \suc {\alpha'} {u'} \qquad
      u = u' \in \NAT(\beta)
    }{t = t' \in \NAT(\beta+1)}
\end{gather*}

@@

% \begin{lemma}[PER Subsumption]
% \label{lem:subsump} \hfill
% \begin{enumerate}
% \item If $\alpha \leq \beta$ then $\NAT(\alpha) \subseteq \NAT(\beta)$.
% \item If $\F(\alpha) \leq \F'(\alpha)$ for all $\alpha \in \SIZE$, then $\FORALL \F \subseteq \FORALL \F'$.
% \item IF $\A' \subseteq \A$ and $\F(u) \subseteq \F'(u)$ for all $u \in \A'$, then $\PIAF \subseteq \PI\,\A'\,\F'$.
% \end{enumerate}
% \end{lemma}

@@

By induction on $\ell \in \NN$ we define the PER family
$\_ = \_ \in \Set[\ell]$ of types
together with the extension $\EL[\ell] T$ (for $T = T' \in \Set[\ell]$)
which is PER of values of type $A$.
The rules for \fbox{$T = T' \in \SET[\ell]$} are:
\[
\begin{array}{l@{\qquad}l}
  \ru{% T \evalsto \up{\Set[\ell']} n \qquad
      % T' \evalsto \up{\Set[\ell'']} n' \qquad
      % n = n' \in \NE
      T = T' \in \NEE
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \NEE
\\[3ex]
  \ru{T \evalsto \Nat\,\alpha \qquad T' \evalsto \Nat\,\alpha
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \NAT(\alpha)
\\[3ex]
  \rux{T \evalsto \Set[\ell'] \qquad T' \evalsto \Set[\ell']
     }{T = T' \in \SET[\ell]
     }{\ell'{<}\ell}
  & \EL[\ell](T) = \SET[\ell']
\\[3ex]
  \rul{T \evalsto \piT A B \qquad
      T' \evalsto \piT{A'}{B'} \\
      A = A' \in \SET[\ell] \qquad
      B[u] = B'[u'] \in \SET[\ell] \mforall u = u' \in \EL[\ell](A)
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \PI(\EL[\ell](A),\;u \mapsto \EL[\ell](B[u]))
\\[3ex]
  \rul{T \evalsto \piT \Size B \quad
      T' \evalsto \piT{\Size}{B'} \quad
      B[\alpha] = B'[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \PI(\SIZE,\; \alpha \mapsto \EL[\ell](B[\alpha]))
\\[3ex]
  \rul{T \evalsto \forallT B \quad
      T' \evalsto \forallT{B'} \quad
      B[\alpha] = B'[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \FORALL(\alpha \mapsto \EL[\ell](B[\alpha]))
\end{array}
\]
All relations involved here are closed under weak head expansion.
\begin{lemma}[Well-definedness]
  Let $\DD :: T_1 = T_2 \in \SET[\ell]$.
  \begin{enumerate}
  \item Symmetry: $T_2 = T_1 \in \SET[\ell]$.
  \item Transitivity: If $T_2 = T_3 \in \SET[\ell]$ then $T_1 = T_3 \in \SET[\ell]$.
  \item Extension: $\EL[\ell](T_1) = \EL[\ell](T_2)$ and ``both'' are PERs.
  \end{enumerate}
\end{lemma}
\begin{proof}
  Simultaneously by induction on $\DD$.
\end{proof}

@@

\subsection{Subtyping}
\label{sec:subty}

The semantic types (PERs) admit subsumption:

\begin{lemma}[Subsumption]
\label{lem:subsump} \hfill
\begin{enumerate}
\item \label{it:natsub} If $\alpha \leq \beta$ then $\NAT(\alpha) \subseteq \NAT(\beta)$.
\item \label{it:allsub} If $\F(\alpha) \leq \F'(\alpha)$ for all $\alpha \in \SIZE$, then $\FORALL \F \subseteq \FORALL \F'$.
\item \label{it:pisub}  If $\A' \subseteq \A$ and $\F(u) \subseteq \F'(u)$ for all $u \in \A'$, then $\PIAF \subseteq \PI\,\A'\,\F'$.
\item \label{it:usub}  If $\ell \leq \ell'$ then $\SET[\ell] \subseteq \SET[\ell']$.
\end{enumerate}
\end{lemma}
\begin{proof}
  Propositions (\ref{it:natsub}--\ref{it:pisub}) are clear.
  For (\ref{it:usub}), prove $T = T' \in \SET[\ell']$ by induction on $T = T' \in \SET[\ell]$.
  For the base types this is direct, let us look at the function space.
\[
  \rul{T \evalsto \piT A B \qquad
      T' \evalsto \piT{A'}{B'} \qquad
      A = A' \in \SET[\ell] \qquad
      B[u] = B'[u'] \in \SET[\ell] \mforall u = u' \in \EL[\ell](A)
    }{T = T' \in \SET[\ell]}
\]
  By induction hypothesis, $A = A' \in \SET[\ell']$, and we have $\EL[\ell'](A) = \EL[\ell](A)$ by Lemma~\ref{lem:indep}.  Assuming $u = u' \in \EL(A)$, we get $B[u] = B'[u'] \in \SET[\ell']$ by induction hypothesis on $B[u] = B'[u'] \in \SET[\ell]$.
\end{proof}
We can define subtyping of type values \fbox{$T \leq T'$} by induction on $T \in \SET[\ell]$ and $T' \in \SET[\ell']$.
Simultaneously, we need to prove correctness, namely that $T \leq T'$ implies $\EL(T) \subseteq \EL(T')$.  The correctness follows from Lemma~\ref{lem:subsump} and we do not spell it out here.
\begin{gather*}
  \ru{T = T' \in \NEE
    }{T \leq T'}
\qquad
  \ru{T \evalsto \Nat \alpha \qquad
      T' \evalsto \Nat \alpha' \qquad
      \alpha \leq \alpha'
    }{T \leq T'}
\qquad
  \ru{T \evalsto \Set[\ell_0] \qquad
      T' \evalsto \Set[\ell_0'] \qquad
      \ell_0 \leq \ell_0'
    }{T \leq T'}
\\[2ex]
  \ru{T \evalsto \piT A B \qquad
      T' \evalsto \piT {A'}{B'} \qquad
      A' \leq A \qquad
      B[u] \leq B'[u'] \mforall u = u' \in A'
    }{T \leq T'}
\\[2ex]
  \ru{T \evalsto \epiT \Size B \qquad
      T' \evalsto \epiT {\Size}{B'} \qquad
      B[\alpha] \leq B'[\alpha] \mforall \alpha \in \SIZE
    }{T \leq T'}
\end{gather*}
\begin{lemma}[Subtyping is preorder]
\label{lem:preord}
\bla
\begin{enumerate}
\item If\/ $T = T' \in \SET[\ell]$ then $T \leq T'$.
\item If\/ $T_1 \leq T_2$ and $T_2 \leq T_3$ then $T_1 \leq T_3$.
\end{enumerate}
\end{lemma}

% \begin{lemma}[Universe subsumption]
%   \label{lem:usubsump}
%   If $\ell \leq \ell'$ then $\SET[\ell] \subseteq \SET[\ell']$.
% \end{lemma}
% \begin{proof}
%   Prove $T = T' \in \SET[\ell']$ by induction on $T = T' \in \SET[\ell]$.
%   For the base types this is direct, let us look at the function space.
% \[
%   \rul{T \evalsto \piT A B \qquad
%       T' \evalsto \piT{A'}{B'} \qquad
%       A = A' \in \SET[\ell] \qquad
%       B[u] = B'[u'] \in \SET[\ell] \mforall u = u' \in \EL[\ell](A)
%     }{T = T' \in \SET[\ell]}
% \]
%   By induction hypothesis, $A = A' \in \SET[\ell']$, and we have $\EL[\ell'](A) = \EL[\ell](A)$ by Lemma~\ref{lem:indep}.  Assuming $u = u' \in \EL(A)$, we get $B[u] = B'[u'] \in \SET[\ell']$ by induction hypothesis on $B[u] = B'[u'] \in \SET[\ell]$.
% \end{proof}

\subsection{Type shapes}
\label{sec:shape}
